---
title: "hw02-Explore Gapminder and use dplyr"
author: JennyHuang
date: 2018-09-20
output: github_document
---
```{r setup, include=FALSE}
chooseCRANmirror(graphics=FALSE, ind=1)
knitr::opts_chunk$set(echo = TRUE)
```

## Load the data
```{r}
install.packages("gapminder")
install.packages("tidyverse")
library(gapminder)
library(tidyverse)
```

## Smell test the data
Before exploring the dataset in detail, I will take a look at the head and tail of the dataset.
```{r}
head(gapminder) ## this allows us to see the top six rows of the data

tail(gapminder) ## this allows us to see the bottom six rows of the data
```

To determine whether 'gapminder' is a list, a vector, a matrix, or a dataframe, we can use `str()` function: 
```{r}
str(gapminder) # this tells use that gapminder is a tibble, a dataframe that make working in the tidyverse easier. 
```
`str()` function not only tells the type/class of gapminder, it also shows that this dataset contains 1704 observations and 6 variables. 

We can also use `class()` to explore the type of gapminder. 
```{r}
class(gapminder)
```
`class() `function also confirms that gapminder is a dataframe!

Although `str()` already tells number of columns and rows in gapminder, there are different commands we can use to check it. 
```{r}
ncol(gapminder) ## total columns in the dataset
nrow(gapminder)  ## total rows in the dataset 
```
To check data type of each variable, we can also use `str()`
```{r}
str(gapminder)
```
From `str()` output, we can see that:

-  country is factor
-  continent is factor
-  year is integer 
-  lifeExp is number
-  population is integer

## Explore individual variables
First, I will select variables I want to explore from gapminder
```{r}
life=gapminder$lifeExp ## this is a quantitative variable
continent=gapminder$continent ## this is a categorical variable
```

There are different ways to check the range of individual variables. The `range()`, `min()`, `max()` function tells us the possible values of life expectancy. 
```{r}
range(life) ## gives the range of life expendence 
min(life)
max(life)
```
For categorical variable, we can use `table()` function to check the count value under each categories.
```{r}
table(continent) ## example of a categorical variable
```

## Explore various plot types
First, I want to explore the relationship between life expectancy and gdp per capita in Americas.
```{r lifeExp vs. gdpPercap in Americas}
gapminder %>% 
  filter(continent=='Americas') %>% 
  ggplot(aes(lifeExp, gdpPercap))+geom_point()
```

Noticed the relationship between life expectancy and gdp per capita in Americas, now I want to explore life expectancy along, see how the data is distributed. Let's explore this within Asia countries this time
```{r Histogram of lifeExp in Aisa}
gapminder %>% 
  filter(continent=='Asia') %>% 
  ggplot( aes(lifeExp))+
  geom_histogram(bins=50)
```

What does it look like when life expectancy is between 60 and 80? This time I will combine histogram with density plot:

```{r Asia btw 60-80}
gapminder %>% 
  filter(continent=='Asia') %>% 
  filter(lifeExp<='80'&lifeExp>='60') %>% 
  ggplot( aes(lifeExp)) +
  geom_histogram(aes(y=..density.., fill="orange"))+
  geom_density()
```

How does the life expectancy varies among continents? One way to explore this is to use boxplot
```{r boxplot options}
gapminder %>% 
  ggplot(aes(continent,lifeExp, fill=continent)) +
  geom_boxplot()+
  scale_fill_brewer(palette = "Pastel2")
```

Another way to look at this is using the combination of violin and jitter. This will maximize the presentation of actual data.
```{r violin and jitter}
gapminder %>% 
  ggplot(aes(continent,lifeExp, fill=continent)) +
geom_violin()+
  geom_jitter(alpha =0.2)+
  scale_fill_brewer(palette = "Pastel2")+
  guides(fill=FALSE) ## Noticed from bove boxplot, we propobaly don't need a legend since it already show color for each continent
```

## Use filter(), select() and %>%

Now I will use both `filter()` and `select()`, and piping into `ggplot()`. This time let explore the relationship between life expectancy range from 50 to 80 and gdp per capita in a different scale.
```{r lifeExp vs. gdp in Asia and Europe}
gapminder %>% 
  select(continent,lifeExp,gdpPercap) %>% 
  filter(continent=='Europe'|continent=='Asia') %>% 
  filter(lifeExp<='80'&lifeExp>='50') %>% 
  ggplot(aes(lifeExp, gdpPercap, color=continent)) +
  geom_point()+
  scale_y_log10() +
  geom_smooth(method = "lm") +
  scale_fill_brewer(palette = "Pastel2")
```

## But I want to do more!

Evaluate this code and describe the result. Presumably the analyst’s intent was to get the data for Rwanda and Afghanistan.
```{r}
filter(gapminder, country == c("Rwanda", "Afghanistan"))
```
Above code showed that there are 6 records for each of Rwanda and Afghanistan. Let validate it with Rwanda data alone
```{r}
filter(gapminder, country == "Rwanda")
```
From above code, clearly there are more than 6 records for Rwanda, that means the code provided was incorrect because it lead to miss data.

Here are two ways to correct this:
```{r}
filter(gapminder, country %in% c("Rwanda", "Afghanistan"))
filter(gapminder, country == "Rwanda" | country == "Afghanistan")
```

Present numerical tables in a more attractive form, such as using knitr::kable().
```{r}
filter(gapminder, country %in% c("Rwanda", "Afghanistan")) %>% 
knitr::kable()
```

Use more of the dplyr functions for operating on a single table.

Using above table as an example, I will add a column call GDP that contains the GDP value for each countries.
```{r}
gapminder%>% 
  filter(country %in% c("Rwanda", "Afghanistan")) %>% 
  mutate(GPD=pop*gdpPercap)
```

